```{r echo=FALSE, results='hide'}
creation <- paste( Sys.time(), Sys.timezone() )
options( scipen = 999 )    ## remove scientific notation
setwd("C:\\Users\\Alex\\Documents\\Coursera\\Johns Hopkins\\Data Science - Specialization\\005 - Reproducible Research\\OTHERS\\Coursera.ReproducibleResearch.Code\\Proj1")
```

# Reproducible Research: Peer Assessment 1
Generated via knir @ `r creation`

## Loading, describing and pre-processing the data

### Load the data from *activity.csv*

```{r}
activity.df <- read.csv( "activity.csv", colClasses = c("integer", "Date", "factor"), stringsAsFactors = FALSE )
```

### Describing the dataset

Sneak peek of the data:
```{r}
head(activity.df)
```

Number of observations:
```{r}
dim(activity.df)
```

Number of dimensions:
```{r}
colnames(activity.df)
```

Verify the existency of NAs (missing data) per column:
```{r}
colSums( is.na( activity.df ) )
```

Overview of the data:
```{r}
summary(activity.df)
```

### Pre-process/transform the data

Remove NAs (missing data):
```{r}
activity.noNAs.df <- na.omit(activity.df)
#rownames(activity.noNAs.df) <- NULL
```

Sneak peek of the data without NAs (missing values):
```{r}
head(activity.noNAs.df)
```

Number of observations (originally `r nrow( activity.df )` rows):
```{r }
dim(activity.noNAs.df)
```

Overview of the data:
```{r }
summary(activity.noNAs.df)
```




## What is [mean total number] of [steps] taken [per day]?
For this part, the [missing values] in the dataset will be [ignored].

Calculate the [total number of steps] taken [per day]:
```{r}
total.steps.per_day <- tapply( activity.noNAs.df$steps, activity.noNAs.df$date, sum )
```

Histogram of the [total number of steps] taken [each day]:
```{r}
hist( total.steps.per_day
    , breaks = 15
    , xlab="Steps per day"
    , ylab="Frequency"
    , ylim = c( 0, 20 )
    , main="[Total number] of [steps] taken [per day]"
)
rug( total.steps.per_day )
```

[Mean] of the [total number of steps] taken [per day]:
```{r}
steps.mean.per_day <- mean( total.steps.per_day )
steps.mean.per_day
```

[Median] of the [total number of steps] taken [per day]:
```{r}
median.steps.per_day <- median( total.steps.per_day )
median.steps.per_day
```





## What is the [average] [daily activity] pattern?

Make a [time series] plot (i.e. type = "l") of the [5-minute] interval (x-axis) and the [average number of steps] taken, [averaged across] all days (y-axis).

```{r}
avg.steps.df <- aggregate( activity.noNAs.df$steps
                         , list( interval = as.numeric( as.character( activity.noNAs.df$interval ) ) )
                         , FUN = mean 
                         )
colnames( avg.steps.df ) <- c("interval","steps.mean")
plot( avg.steps.df , type = "l"
    , xlab="[5-minute] interval"
    , ylab="Steps"
    , main = "[Mean] number of [Steps] by [Interval]"
    )
```

Which [5-minute interval], on [average across all the days] in the dataset, contains the [maximum number of steps]?

```{r cleandata_maxsteps}
avg.steps.df[ avg.steps.df$steps.mean== max( avg.steps.df$steps.mean) , ]
```



## Imputing missing values

The presence of [missing data] values may introduce bias into some calculations or summaries of the data.

Calculate and report the [total number of missing values] in the dataset (i.e. the total number of rows with `NA`s):
```{r}
colSums( is.na( activity.df ) )
```

The work-around is the imputation of the [mean for that 5-minutes interval] to replace all [NA] value in the [steps column].

Create a new dataset that is equal to the original dataset but with the [missing data filled] in
```{r new_dataset_filled}
imputation.df <- activity.df 
for( idx in 1:nrow( imputation.df ) ) {
  if( is.na( imputation.df$steps[ idx ] ) ) {
     imputation.df$steps[ idx ] <- avg.steps.df$steps.mean[ avg.steps.df$interval==imputation.df$interval[ idx ] ]
  }
}
new.steps.per_day <- tapply( imputation.df$steps, imputation.df$date, sum )
```


Make a histogram of the [total number of steps] taken [each day].
```{r}
par( mfrow=c(1,2) )
hist( total.steps.per_day
    , breaks = 15
    , xlab="Steps per day"
    , ylab="Frequency"
    , ylim = c( 0, 20 )
    , main="[Total number] of [steps]\ntaken [per day]"
)
rug( total.steps.per_day )
hist( new.steps.per_day
    , breaks = 15
    , xlab="Steps per day"
    , ylab="Frequency"
    , ylim = c( 0, 20 )
    , main="Imputation of NAs"
)
rug(new.steps.per_day)

```

After recalculate [mean] and [median] of the [total number of steps] taken [per day] (imputation of missing data),
verify what is the [impact] of the [imputation of missing data] on the estimates of the [total daily number of steps]?

```{r}
newtot.steps <- tapply( imputation.df$steps , imputation.df$date , FUN = sum )
newmean <- mean( newtot.steps )
newmedian <- median( newtot.steps )
data.frame( Mean=c(steps.mean.per_day, newmean, abs(steps.mean.per_day-newmean))
          , Median=c(median.steps.per_day , newmedian , abs(median.steps.per_day-newmedian))
          , row.names=c("Original","Imputation", "Difference" )
          )
```

**There is no change to the [mean], however the new [median] calculated with [imputation of the missing data] is higher.**

## Are there differences in activity patterns between [weekdays] and [weekends]?

Create a new factor variable in the dataset with two levels: [weekday] and [weekend] indicating whether a given [date] is a [week day] or [weekend day]:
```{r}
imputation.df$weekdays <- factor( format( imputation.df$date , "%A" ) )
levels(imputation.df$weekdays) <- list( weekday = c("Monday", "Tuesday"
                                                   , "Wednesday", "Thursday"
                                                   , "Friday"
                                                   )
                                      , weekend = c("Saturday", "Sunday" ) 
                                      )
table(imputation.df$weekdays)
```

Panel plot containing a [time series plot] (i.e. type = "l") of the [5-minute interval] (x-axis) and the [average number of steps] taken, [averaged across] all [weekday days] or [weekend days] (y-axis). 
```{r}
library( ggplot2 )
newavg.steps.df <- aggregate( imputation.df$steps
                         , list( interval = as.numeric( as.character( imputation.df$interval ) )
                               , weekdays = imputation.df$weekdays )
                         , FUN = mean )
colnames( newavg.steps.df )[3] <-"steps.mean"

ggplot( newavg.steps.df, aes( interval, steps.mean ) ) +
        geom_line( color = "blue", size = 0.7)         +
        facet_wrap( ~ weekdays , ncol = 1 )            +
        labs( title = "Time series plot of the [5-minute interval]"
            , x = "Interval"
            , y = "Number of steps" )
```
